---
title: "Roulette Simulation and Statistics"
output: rmarkdown::github_document
---

## Simulation Setup

Starting by randomly selecting each color (with appropriate probability) to simulate 1000 spins of the wheel. Starting balance is \$100 and betting increment is \$5. The below table has all info used in the following few graphs.

```{r}
suppressWarnings(library(tidyverse))

bet_increment = 5
begn_balance = 100

spins = data.frame(
spin_number = paste0(seq(1:1000))) %>% 
mutate(color = sample(c("Black","Red","Green"),
                       nrow(.),
                       prob=c(.4734,.4734,.0532),
                       replace=TRUE)
       ) %>% 
mutate(bet = bet_increment,
       all_green = if_else(color == "Green",bet*17,-bet),
       all_red = if_else(color == "Red",bet,-bet),
       all_black = if_else(color == "Black",bet,-bet),
       green_balance = begn_balance+cumsum(all_green),
       red_balance = begn_balance+cumsum(all_red),
       black_balance = begn_balance+cumsum(all_black),
       spin_number = as.integer(spin_number),
       streak=sequence(rle(color)$lengths) - 1) %>% 
  group_by(color) %>% mutate(freq = n())

spins %>% head(10)
```

To ensure the random sampling is valid, the below graph shows the distribution of spins by color. This looks to be as expected.

```{r}
ggplot(spins %>% distinct(color,freq)) +
  aes(x = reorder(color,-freq), y = freq, fill = color) +
  geom_col() +
  scale_fill_manual(
    values = c(Black = "black",
    Green = "darkgreen",
    Red = "firebrick")
  ) +
  labs(
    x = "Color",
    y = "Frequency",
    title = "Spin Distribution"
  ) +
  theme_minimal() +
   guides(fill="none")
```

A quick check of the consecutive spins for each color is as follows. Red and black decay at an expected rate for a random event.

```{r}
spins %>% 
  group_by(color,streak) %>% 
  summarise(spins = n()) %>% 
  pivot_wider(names_from = color,values_from = spins,values_fill = 0) %>% 
  janitor::adorn_percentages("col") %>% as_tibble() %>% round(2) %>% filter(streak < 6)
```

### Simulation 1 - Constant Amount

The first simulation considers a constant bet for each spin. Here is \$5 for every spin for black and green.

```{r}
ggplot(spins %>% select(spin_number,green_balance,red_balance,black_balance) %>%  pivot_longer(cols = contains("balance"))) +
  aes(x = spin_number, y = value,group=name,color=name) +
  geom_line() +
    scale_color_manual(
    values = c(`green_balance` = "darkgreen",
               `red_balance` = "firebrick",
               `black_balance` = "black")) +
  theme_minimal() +
  geom_line(linewidth =1) +
  scale_y_continuous(label=scales::dollar) +
  guides(color="none") +
  labs(
    x = "Spins",
    y = "Balance",
    title = "$5 Every Spin - All Colors"
  )
```

### Simulation 2 - Martingale

Martingale increases bet amounts after each loss where a win would offset all previous losses.

First, I create the doubling sequence.
```{r}
pascalTriangle <- function(h) {
    lapply(0:h, function(i) choose(i, 0:i))
}

base_bet = 4

triangle = unlist(lapply(pascalTriangle(25), sum)) %>% as.data.frame() %>% mutate(row_number()) %>% 
  rename(factor_increase = 1,spin =2) %>% 
  mutate(bet_amount = base_bet*factor_increase) %>% select(spin,bet_amount)

triangle %>% head(10)
```

I'll simulate 100 spins of the wheel with a beginning balance of $200 and a $4 baseline bet (2% of total - the 2% ratio seems to be the best balance of risk/reward).
```{r}
begn_balance = 200

spins = data.frame(
spin_number = paste0(seq(1:100))) %>% 
mutate(color = sample(c("Black","Red","Green"),
                       nrow(.),
                       prob=c(.4734,.4734,.0532),
                       replace=TRUE)
       ) %>% 
mutate(streak = coalesce(case_when(color == "Red" & lag(color) == "Red" ~ 1,
                          color == "Red" & lag(color) != "Red" ~ lag(sequence(rle(color != "Red")$lengths)+1),
                 color != "Red" ~ sequence(rle(color != "Red")$lengths)),1)
       ) %>% 
  left_join(triangle,by=c("streak"="spin")) %>% 
  rename(red_bet = bet_amount) %>% 
  mutate(PL = if_else(color == "Red",red_bet,-red_bet),
       balance = begn_balance+cumsum(PL))

 
spins %>% head(10)
```

```{r}
ggplot(spins) +
    aes(x = as.integer(spin_number), y = balance) +
    geom_line(color = "firebrick",linewidth =1) +
    theme_minimal() +
  scale_y_continuous(label=scales::dollar) +
  scale_x_continuous(label=scales::comma) +
  labs(
    x = "Spins",
    y = "Balance",
    title = "Martingale Method",
    subtitle = "Double Bet on Every Losing Spin"
  )
```

Show summary stats for bets, running balance, and single spin profit/loss
```{r}
spins %>% 
  summarise(min_bet = min(red_bet),max_bet = max(red_bet),min_balance = min(balance),min_PL = min(PL),max_PL = max(PL)) %>% 
  mutate(max_drawdown = begn_balance - min_balance) %>% pivot_longer(cols = everything())
```
### Simulation 3 - Fibonacci

Create Fibonacci sequence; each number is the sum of the two preceding values.
```{r}
base_bet = 5

n <- 15
fib <- numeric(n)
fib[1] <- base_bet
fib[2] <- base_bet
for (i in 3:n)
{
  fib[i] <- fib[i-1]+fib[i-2] 
}
fib = data.frame(bet_amount  = fib,spin = seq(1:n)) %>% 
  mutate(rolling_fib_amt = cumsum(bet_amount))

fib
```

```{r}
begn_balance = 200

spins = data.frame(
spin_number = paste0(seq(1:100))) %>% 
mutate(color = sample(c("Black","Red","Green"),
                       nrow(.),
                       prob=c(.4734,.4734,.0532),
                       replace=TRUE)
       ) %>% 
mutate(streak = coalesce(case_when(color == "Red" & lag(color) == "Red" ~ 1,
                          color == "Red" & lag(color) != "Red" ~ lag(sequence(rle(color != "Red")$lengths)+1),
                 color != "Red" ~ sequence(rle(color != "Red")$lengths)),1)
       ) %>% 
  left_join(fib,by=c("streak"="spin")) %>% 
  rename(red_bet = bet_amount) %>% 
  mutate(PL = if_else(color == "Red",red_bet,-red_bet),
       balance = begn_balance+cumsum(PL))

spins %>% head(10)
```


```{r}
ggplot(spins) +
    aes(x = as.integer(spin_number), y = balance) +
    geom_line(color = "firebrick",linewidth =1) +
    theme_minimal() +
  scale_y_continuous(label=scales::dollar) +
  scale_x_continuous(label=scales::comma) +
  labs(
    x = "Spins",
    y = "Balance",
    title = "Fibonacci Method",
    subtitle = "Add Previous Two Bet Values on Every Losing Spin"
  )
```

### Simulation 4 - Consecutive Color Contrarian
If the spins are truly random, any run of more than two (zero being the beginning of a new trend), has very low odds. Below table simulates betting the opposite color on streaks greater than 1. No action is taken when green appears more than once.

This method involves taking a betting increment and allocating 90% to the inverse color and 10% to green as "insurance". In this case I use a $10 increment; $9 is allotted to the inverse color of the streak, and $1 is allocated to green. I made this simulation for 1,000 spins instead of 100 considered bets aren't taking place on most spins.
```{r}
increment = 10

spins = data.frame(
spin_number = paste0(seq(1:1000))) %>% 
mutate(color = sample(c("Black","Red","Green"),
                       nrow(.),
                       prob=c(.4734,.4734,.0532),
                       replace=TRUE)
       ) %>% 
mutate(streak=sequence(rle(color)$lengths) - 1) %>% 
ungroup() %>% 
mutate(prev_run = lag(streak),
       PL = case_when(lag(color) != color & 
                        lag(color) != "Green" & 
                        color != "Green" &
                        prev_run >= 2 ~ increment - (increment *0.1),
                      
                      lag(color) != color & color == "Green" & prev_run >= 2 ~ ((increment*0.1)*17)-(increment - (increment *0.1)),
                      lag(color) == color & prev_run >= 2 ~ -increment,
                      prev_run < 2 | lag(color) == "Green" ~ 0)) %>% 
  replace(is.na(.), 0) %>% 
  mutate(balance = begn_balance+cumsum(PL))

spins %>% head(10)
```


```{r}
ggplot(spins) +
    aes(x = as.integer(spin_number), y = balance) +
    geom_line(color = "navy",linewidth =1) +
    theme_minimal() +
  scale_y_continuous(label=scales::dollar) +
  scale_x_continuous(label=scales::comma) +
  labs(
    x = "Spins",
    y = "Balance",
    title = "Color Contrarian Method",
    subtitle = "Bet against the current streak on opposite color"
  )
```