---
title: "Roulette Simulation and Statistics"
output: rmarkdown::github_document
---

## Simulation Setup

Starting by randomly selecting each color (with appropriate probability) to simulate 5000 spins of the wheel. Starting balance is \$100 and betting increment is \$5. The below table has all info used in the following few graphs.

```{r}
suppressMessages(library(tidyverse))

bet_increment = 5
begn_balance = 100

spins = data.frame(
spin_number = paste0(seq(1:5000))) %>% 
mutate(color = sample(c("Black","Red","Green"),
                       nrow(.),
                       prob=c(.4734,.4734,.0532),
                       replace=TRUE)
       ) %>% 
mutate(bet = bet_increment,
       all_green = if_else(color == "Green",bet*17,-bet),
       all_red = if_else(color == "Red",bet,-bet),
       green_balance = begn_balance+cumsum(all_green),
       red_balance = begn_balance+cumsum(all_red),
       spin_number = as.integer(spin_number),
       streak=sequence(rle(color)$lengths) - 1) %>% 
  group_by(color) %>% mutate(freq = n())

spins %>% head(10)
```

To ensure the random sampling is correct, the below graph shows the distribution of spins by color. This looks to be as expected.

```{r}
ggplot(spins %>% distinct(color,freq)) +
  aes(x = reorder(color,-freq), y = freq, fill = color) +
  geom_col() +
  scale_fill_manual(
    values = c(Black = "black",
    Green = "darkgreen",
    Red = "firebrick")
  ) +
  labs(
    x = "Color",
    y = "Frequency",
    title = "Spin Distribution"
  ) +
  theme_minimal() +
   guides(fill="none")
```

A quick check of the consecutive spins for each color is as follows. Red and black decay at an expected rate for a random event, while the odds of the wheel landing on 2 greens in a row is \< 5%

```{r}
spins %>% 
  group_by(color,streak) %>% 
  summarise(spins = n()) %>% 
  pivot_wider(names_from = color,values_from = spins,values_fill = 0) %>% 
  janitor::adorn_percentages("col") %>% as_tibble() %>% round(2) %>% filter(streak < 6)
```

### Simulation 1 - Constant Amount

The first simulation considers a constant bet for each spin. Here is \$5 for every spin for black and green.

```{r}
ggplot(spins %>% select(spin_number,green_balance,red_balance) %>%  pivot_longer(cols = contains("balance"))) +
  aes(x = spin_number, y = value,group=name,color=name) +
  geom_line() +
    scale_color_manual(
    values = c(`green_balance` = "darkgreen",
               `red_balance` = "firebrick")) +
  theme_minimal() +
  geom_line(linewidth =1) +
  scale_y_continuous(label=scales::dollar) +
  guides(color="none") +
  labs(
    x = "Spins",
    y = "Balance",
    title = "$5 Every Spin"
  )
```

### Simulation 2 - Martingale

Martingale increases bet amounts after each loss where a win would offset all previous losses.

Create doubling sequence...

```{r}
pascalTriangle <- function(h) {
    lapply(0:h, function(i) choose(i, 0:i))
}

base_bet = 5

triangle = unlist(lapply(pascalTriangle(25), sum)) %>% as.data.frame() %>% mutate(row_number()) %>% 
  rename(factor_increase = 1,spin =2) %>% 
  mutate(bet_amount = base_bet*factor_increase) %>% select(spin,bet_amount)

triangle %>% head(10)
```

I'll simulate 500 spins of the wheel with the same beginning balance and betting increment as before.

```{r}
begn_balance = 100

spins = data.frame(
spin_number = paste0(seq(1:500))) %>% 
mutate(color = sample(c("Black","Red","Green"),
                       nrow(.),
                       prob=c(.4734,.4734,.0532),
                       replace=TRUE)
       ) %>% 
mutate(red_streak = coalesce(case_when(color == "Red" & lag(color) == "Red" ~ 1,
                          color == "Red" & lag(color) != "Red" ~ lag(sequence(rle(color != "Red")$lengths)+1),
                 color != "Red" ~ sequence(rle(color != "Red")$lengths)),1)
       ) %>% 
  left_join(triangle,by=c("red_streak"="spin")) %>% 
  rename(red_bet = bet_amount) %>% 
  mutate(all_red = if_else(color == "Red",red_bet,-red_bet),
       red_balance = begn_balance+cumsum(all_red))

 
spins %>% head(10)
```

```{r}
ggplot(spins) +
    aes(x = as.integer(spin_number), y = red_balance) +
    geom_line(color = "firebrick",linewidth =1) +
    theme_minimal() +
  scale_y_continuous(label=scales::dollar) +
  scale_x_continuous(label=scales::comma) +
  labs(
    x = "Spins",
    y = "Balance",
    title = "Martingale Method",
    subtitle = "Double Bet on Every Losing Spin"
  )
```

```{r}
summary(spins %>% select(red_bet,red_balance))
```
### Simulation 3 - Fibonacci


Create Fibonacci sequence; each number is the sum of the two preceding values.
```{r}
base_bet = 5

n <- 15
fib <- numeric(n)
fib[1] <- base_bet
fib[2] <- base_bet
for (i in 3:n)
{
  fib[i] <- fib[i-1]+fib[i-2] 
}
fib = data.frame(bet_amount  = fib,spin = seq(1:n))

fib
```

```{r}
begn_balance = 100

spins = data.frame(
spin_number = paste0(seq(1:500))) %>% 
mutate(color = sample(c("Black","Red","Green"),
                       nrow(.),
                       prob=c(.4734,.4734,.0532),
                       replace=TRUE)
       ) %>% 
mutate(red_streak = coalesce(case_when(color == "Red" & lag(color) == "Red" ~ 1,
                          color == "Red" & lag(color) != "Red" ~ lag(sequence(rle(color != "Red")$lengths)+1),
                 color != "Red" ~ sequence(rle(color != "Red")$lengths)),1)
       ) %>% 
  left_join(fib,by=c("red_streak"="spin")) %>% 
  rename(red_bet = bet_amount) %>% 
  mutate(all_red = if_else(color == "Red",red_bet,-red_bet),
       red_balance = begn_balance+cumsum(all_red))

 
spins
```


```{r}
ggplot(spins) +
    aes(x = as.integer(spin_number), y = red_balance) +
    geom_line(color = "firebrick",linewidth =1) +
    theme_minimal() +
  scale_y_continuous(label=scales::dollar) +
  scale_x_continuous(label=scales::comma) +
  labs(
    x = "Spins",
    y = "Balance",
    title = "Fibonacci Method",
    subtitle = "Add Previous Two Bet Values on Every Losing Spin"
  )
```

### Simulation 4 - Consecutive Color Contrarian
If the spins are truly random, any run of more than two (zero being the beginning of a new trend), has very low odds. Next table simulates betting the opposite color on streaks greater than 1. No action is taken when green appears more than once.

This method involves taking a betting increment and allocating 90% to the inverse color and 10% to green as "insurance". In this case I use a $10 increment; $9 is allotted to the inverse color of the streak, and $1 is allocated to green. Simulation is 5,000 spins.
```{r}
increment = 10

spins = data.frame(
spin_number = paste0(seq(1:5000))) %>% 
mutate(color = sample(c("Black","Red","Green"),
                       nrow(.),
                       prob=c(.4734,.4734,.0532),
                       replace=TRUE)
       ) %>% 
mutate(streak=sequence(rle(color)$lengths) - 1) %>% 
ungroup() %>% 
mutate(prev_run = lag(streak),
       PL = case_when(lag(color) != color & 
                        lag(color) != "Green" & 
                        color != "Green" &
                        prev_run >= 1 ~ increment - (increment *0.1),
                      
                      lag(color) != color & color == "Green" & prev_run >= 1 ~ ((increment*0.1)*17)-(increment - (increment *0.1)),
                      lag(color) == color & prev_run >= 1 ~ -increment,
                      prev_run < 1 | lag(color) == "Green" ~ 0)) %>% 
  replace(is.na(.), 0) %>% 
  mutate(balance = begn_balance+cumsum(PL))

spins
```


```{r}
ggplot(spins) +
    aes(x = as.integer(spin_number), y = balance) +
    geom_line(color = "navy",linewidth =1) +
    theme_minimal() +
  scale_y_continuous(label=scales::dollar) +
  scale_x_continuous(label=scales::comma) +
  labs(
    x = "Spins",
    y = "Balance",
    title = "Color Contrarian Method",
    subtitle = "Bet against the current streak on opposite color"
  )
```